---
title: "BM_final"
output: html_document
date: "2024-12-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggplot2)
library(GGally)
```

# load data
```{r}
data = read.csv("Project_2_data.csv")
data
```
# preprocess
```{r}
colnames(data)
data <- data %>%
  drop_na()%>%
  mutate(Main.Stage = case_when(
    X6th.Stage %in% c("IIIA", "IIIC") ~ "III",
    X6th.Stage %in% c("IIA", "IIB") ~ "II",
    X6th.Stage %in% c("IVA", "IVB") ~ "IV",
    TRUE ~ X6th.Stage
  ))%>%
  mutate(
    T.Stage = as.numeric(gsub("T", "", T.Stage)),
    N.Stage = as.numeric(gsub("N", "", N.Stage))
  )%>%
  mutate(
    Race = as.factor(Race),
    Marital.Status = as.factor(Marital.Status),
    Estrogen.Status = as.factor(Estrogen.Status),
    Progesterone.Status = as.factor(Progesterone.Status),
    differentiate = as.factor(differentiate),
    A.Stage = as.factor(A.Stage),
    Status = as.factor(Status),
    Main.Stage = as.factor(Main.Stage)
  )%>%
  select(-X6th.Stage)%>%
  mutate(
    Grade = as.numeric(gsub("[^0-9]", "", Grade)) 
  )%>%
  rename(Regional.Node.Positive = "Reginol.Node.Positive")
data

```

## see the plot after first preprocess
```{r}
for (var in names(data)) {
  # Skip if the column is not numeric
  if (is.numeric(data[[var]])) {
    
    # Histogram
    p1 <- ggplot(data, aes_string(x = var)) +
      geom_histogram(fill = "skyblue", color = "black", bins = 30) +
      labs(title = paste("Histogram of", var), x = var, y = "Frequency") +
      theme_minimal()
    print(p1)
    
    # Boxplot
    p2 <- ggplot(data, aes_string(y = var)) +
      geom_boxplot(fill = "lightgreen", color = "black") +
      labs(title = paste("Boxplot of", var), y = var) +
      theme_minimal()
    print(p2)
  }
}

```
## see the correlation between regional.node.examined and regional.node.positive
```{r}
pairs_data <- data[, c("Regional.Node.Examined", "Regional.Node.Positive")]

pairs(pairs_data, main = "Pairs Plot: Regional Node Examined vs Regional Node Positive",
      pch = 21, bg = "lightblue")

```
Linear Trend: There appears to be a positive association between Regional.Node.Examined and Regional.Node.Positive. As the number of nodes examined increases, the number of positive nodes also tends to increase.
However, the relationship is not perfectly linear; there is noticeable spread in the points.

High Variability: There is significant variability in Regional.Node.Positive values for a given range of Regional.Node.Examined. This suggests that other factors might influence the number of positive nodes beyond the number of examined nodes.

Outliers: A few observations stand out as potential outliers, particularly where Regional.Node.Examined is high, but the number of Regional.Node.Positive remains low (or vice versa). These outliers could be influential points worth further investigation.

## introduce a new variable proportion
```{r}
data = data%>%
  mutate(Node.Proportion = Regional.Node.Positive / Regional.Node.Examined)
```

## see the plot after the second preprocess
```{r}
for (var in names(data)) {
  # Skip if the column is not numeric
  if (is.numeric(data[[var]])) {
    
    # Histogram
    p1 <- ggplot(data, aes_string(x = var)) +
      geom_histogram(fill = "skyblue", color = "black", bins = 30) +
      labs(title = paste("Histogram of", var), x = var, y = "Frequency") +
      theme_minimal()
    print(p1)
    
    # Boxplot
    p2 <- ggplot(data, aes_string(y = var)) +
      geom_boxplot(fill = "lightgreen", color = "black") +
      labs(title = paste("Boxplot of", var), y = var) +
      theme_minimal()
    print(p2)
  }
}
```

## skewness of Tumor.size
```{r}
hist(log(data$Tumor.Size))
hist(sqrt(data$Tumor.Size))
```

## log transformation for tumor.size

```{r}
data = data %>%
  mutate(Tumor.Size = log(Tumor.Size))
```
## see the plot after the third preprocess
```{r}
for (var in names(data)) {
  # Skip if the column is not numeric
  if (is.numeric(data[[var]])) {
    
    # Histogram
    p1 <- ggplot(data, aes_string(x = var)) +
      geom_histogram(fill = "skyblue", color = "black", bins = 30) +
      labs(title = paste("Histogram of", var), x = var, y = "Frequency") +
      theme_minimal()
    print(p1)
    
    # Boxplot
    p2 <- ggplot(data, aes_string(y = var)) +
      geom_boxplot(fill = "lightgreen", color = "black") +
      labs(title = paste("Boxplot of", var), y = var) +
      theme_minimal()
    print(p2)
  }
}
```
```{R, warning = FALSE, echo = FALSE}
#Single hist plot for EDA

#Data frame to long format
df_long <- reshape2::melt(data)

# Create plot
ggplot(df_long, aes(value)) + 
  geom_histogram(binwidth = 0.1, fill = "lightblue", color = "black") + 
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  theme_minimal() +
  labs(title = "Histograms of Variables", x = "Value", y = "Frequency") +
geom_vline(data = df_long, aes(xintercept = ave(value, variable, FUN = mean)), 
           color = "magenta", linetype = "dashed", size = 1, show.legend = FALSE) + 
geom_vline(data = df_long, aes(xintercept = ave(value, variable, FUN = median)), 
           color = "red", linetype = "dotted", size = 1.3, show.legend = FALSE)


```



#Model Construction
#Model1-Baseline Logistic Regression
```{r}
model1 <- glm(Status ~ Age + Race + T.Stage + N.Stage + Tumor.Size + 
                Estrogen.Status + Progesterone.Status, 
              data = data, family = binomial)
summary(model1)
```
#Model2-Logistic Regression with Transformed Variables
```{r}
model2 <- glm(Status ~ Age + Race + Node.Proportion + Tumor.Size + 
                Estrogen.Status + Progesterone.Status, 
              data = data, family = binomial)
summary(model2)
```

#Model3-Interaction Model
```{r}
model3 <- glm(Status ~ Age + Race + Node.Proportion + Tumor.Size + 
                Estrogen.Status * Progesterone.Status, 
              data = data, family = binomial)
summary(model3)
```
#Model4-Stepwise Selection
```{r}
# Remove rows with any missing values before fitting the model
clean_data <- data %>% drop_na()

# Build the full logistic regression model
full_model <- glm(Status ~ ., data = clean_data, family = binomial)

# Perform stepwise selection
model4 <- step(full_model, direction = "both", trace = FALSE)

# Summary of the selected model
summary(model4)

```
#Model5-Fairness: Separate Models by Race
```{r}
# Stratify data by race
data_white <- filter(data, Race == "White")
data_nonwhite <- filter(data, Race != "White")

# White group model
model_white <- glm(Status ~ Age + Node.Proportion + Tumor.Size, 
                   data = data_white, family = binomial)

# Non-white group model
model_nonwhite <- glm(Status ~ Age + Node.Proportion + Tumor.Size, 
                      data = data_nonwhite, family = binomial)

summary(model_white)
summary(model_nonwhite)
```

#Model Validation
```{r}
library(pROC)
library(caret)
# Ensure predictions align with the dataset used for model training
clean_data$predicted_prob <- predict(model4, type = "response")

# ROC Curve for Model 4
roc_curve <- roc(clean_data$Status, clean_data$predicted_prob)
plot(roc_curve, main = "ROC Curve for Model 4", col = "blue")
auc_value <- auc(roc_curve)
print(paste("AUC for Model 4:", round(auc_value, 3)))

# Confusion Matrix (using a cutoff of 0.5)
predicted_class <- ifelse(clean_data$predicted_prob > 0.5, "Dead", "Alive")
conf_matrix <- confusionMatrix(as.factor(predicted_class), as.factor(clean_data$Status), positive = "Dead")
print(conf_matrix)
```

```{r}
# Use cleaned data for residuals and fitted values
clean_data$residuals <- residuals(model4, type = "deviance")
fitted_values <- fitted(model4)

# Residuals vs Fitted Values Plot
library(ggplot2)
ggplot(clean_data, aes(x = fitted_values, y = residuals)) +
  geom_point(color = "blue", alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted Values", x = "Fitted Values", y = "Deviance Residuals") +
  theme_minimal()
```

```{r}
clean_data <- clean_data %>%
  filter(differentiate != "Undifferentiated") %>%
  droplevels()
model4 <- glm(Status ~ Age + Race + T.Stage + N.Stage + Survival.Months +
                Node.Proportion + Regional.Node.Positive + 
                Estrogen.Status + Progesterone.Status,
              data = clean_data, family = binomial)
# Filter out "Undifferentiated" rows in subsets
data_white_filtered <- data_white %>%
  filter(differentiate %in% levels(clean_data$differentiate)) %>%
  droplevels()

data_nonwhite_filtered <- data_nonwhite %>%
  filter(differentiate %in% levels(clean_data$differentiate)) %>%
  droplevels()

# Predict probabilities and evaluate ROC for White group
roc_white <- roc(data_white_filtered$Status,
                 predict(model4, newdata = data_white_filtered, type = "response"))

auc_white <- auc(roc_white)
print(paste("AUC for White group:", round(auc_white, 3)))

# Predict probabilities and evaluate ROC for Non-White group
roc_nonwhite <- roc(data_nonwhite_filtered$Status,
                    predict(model4, newdata = data_nonwhite_filtered, type = "response"))

auc_nonwhite <- auc(roc_nonwhite)
print(paste("AUC for Non-White group:", round(auc_nonwhite, 3)))


```

```{r}
library(broom)

# Extract and format coefficients
coeff_table <- tidy(model4, conf.int = TRUE, exponentiate = TRUE)
significant_coeffs <- coeff_table %>% filter(p.value < 0.05)
print(significant_coeffs)

# Save table as CSV (for your report)
write.csv(significant_coeffs, "significant_model4_coefficients.csv")

```

```{r}
# Boxplot for Node.Proportion by Status
ggplot(data, aes(x = Status, y = Node.Proportion, fill = Status)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Node Proportion by Survival Status", x = "Status", y = "Node Proportion")

# Scatter plot for Survival.Months vs Node.Proportion
ggplot(data, aes(x = Survival.Months, y = Node.Proportion, color = Status)) +
  geom_point(alpha = 0.7) +
  theme_minimal() +
  labs(title = "Survival Months vs Node Proportion", x = "Survival Months", y = "Node Proportion")
```
#Model Validation with sample
#Split Data into Training and Testing Sets
```{r}
set.seed(123)  # For reproducibility
library(caret)

# Create training and testing sets
train_index <- createDataPartition(clean_data$Status, p = 0.8, list = FALSE)
train_data <- clean_data[train_index, ]
test_data <- clean_data[-train_index, ]

# Refit the model using the training set
model4_train <- glm(Status ~ Age + Race + T.Stage + N.Stage + Survival.Months +
                      Node.Proportion + Regional.Node.Positive +
                      Estrogen.Status + Progesterone.Status,
                    data = train_data, family = binomial)

# Predict probabilities on the test set
test_data$predicted_prob <- predict(model4_train, newdata = test_data, type = "response")
```

#Evaluate Model Performance on Test Data
```{r}
# ROC-AUC for test data
library(pROC)
roc_test <- roc(test_data$Status, test_data$predicted_prob)
auc_test <- auc(roc_test)

# Plot ROC curve
plot(roc_test, main = "ROC Curve for Test Data", col = "blue")
print(paste("AUC on Test Data:", round(auc_test, 3)))

# Create confusion matrix using a threshold of 0.5
test_data$predicted_class <- ifelse(test_data$predicted_prob > 0.5, "Dead", "Alive")
conf_matrix_test <- confusionMatrix(as.factor(test_data$predicted_class), 
                                    as.factor(test_data$Status), 
                                    positive = "Dead")
print(conf_matrix_test)
```
#Cross-Validation
```{r}
# Perform k-fold cross-validation (k = 5)
set.seed(123)
cv_control <- trainControl(method = "cv", number = 5, classProbs = TRUE, summaryFunction = twoClassSummary)

cv_model <- train(Status ~ Age + Race + T.Stage + N.Stage + Survival.Months +
                    Node.Proportion + Regional.Node.Positive +
                    Estrogen.Status + Progesterone.Status,
                  data = clean_data, method = "glm", family = "binomial",
                  trControl = cv_control, metric = "ROC")

# Print cross-validation results
print(cv_model)
```

```{r}
threshold <- 0.3
test_data$predicted_class <- ifelse(test_data$predicted_prob > threshold, "Dead", "Alive")
conf_matrix_adj <- confusionMatrix(as.factor(test_data$predicted_class), 
                                   as.factor(test_data$Status), positive = "Dead")
print(conf_matrix_adj)
```

